Mechanising Hylomorphisms for Extraction of Executable Code

### 2025-03-17 paper/paper.tex
Massaged section 3.1

The remainder of the paper is structured as follows:
\begin{itemize}
  \item In Section~\ref{sec:rec-schemes} we give an overview about recursion
schemes.
  \item In Section~\ref{sec:containers} we formalise the type of container
functors ensuring the presence of least and greatest fixed-points for functors
and suitably adapted for program extraction.
  \item In Section~\ref{sec:recursion-schemes} we mechanise folds, unfolds, and
hylomorphisms, as well as proving their uniqueness properties in Coq.
  \item In Section~\ref{sec:examples} we use the framework to formalise examples
of divide-and-conquer, dynamic programming, and mutual recursion algorithms.
Furthermore we verify the short-cut fusion optimisation and show the extracted
optimised code to OCaml.
\end{itemize}

\section{Recursion Schemes}\label{sec:rec-schemes}
The structure of data is very similar to the structure of an algorithm which
processes that data. This relationship manifests in the form of structured
recursion schemes which are widely used in functional languages such Haskell.
Canonical examples are folds (catamorphisms), which consume data, and unfolds
(anamorphisms), which produce it. While some implementations of recursion
schemes like $\operator{foldr}$/$\operator{unfoldr}$ in Haskell are specific to
a particular data structure, in this case $\operator{List}$s, we can generalise
further these ideas to account for generic (co)inductive data types and generic
algorithms operating on them.

@@

\subsection{Mechanising Extractable Container Functors}
\label{sec:containers}

A straightforward mechanisation of functors as defined in
Section~\ref{sec:rec-schemes} does not work in Coq due to the strict positivity
condition. In fact, if we simply require that functors are functions
\coq{F : Type -> Type}
satisfying the necessary identity and composition properties, then
we will not be able to construct least/greatest fixed points of this type.  A
common solution to the strict positivity problem is the use
\emph{containers}~\cite{AbbottAG05}.  A container $S \triangleright P$ is
defined by a type of \emph{shapes} $S : \Type$ and a \emph{family of position
types} indexed by shapes $P : S \to \Type$.  An \emph{extension} of this
container is a functor $\llbracket S \triangleright P \rrbracket$, whose action
on objects is given by
$\llbracket S \triangleright P \rrbracket \; X = \Sigma s : S.\; P\;s \to X$,
and on morphisms is given by post-composition, keeping the shape the same
$\llbracket S \triangleright P \rrbracket \; f = \lambda (s,\;g).\;(s,\;f\circ g)$.
Intuitively, the type of shapes captures the different ``cases'' of an algebraic
datatype, and the positions of a shape are used to identify the elements inside
a container of the given shape.

@@

The straightforward mechanisation of containers and container extensions is
still not enough for our goals of remaining axiom-free, extracting to idiomatic
OCaml code, and avoiding \ocaml{Obj.magic} in the extracted code. The main
problem is that a container extension requires a function from a family of
positions to a type: $\forall s : S.\; P\;s \to X$. On the Coq side, dealing
with the equality of objects that contain (dependent) function types would force
us to add the functional extensionality axiom, and axiom K (or equivalent) to
deal with the heterogeneous equality of position.  For extraction, OCaml's type
system is not powerful enough to represent such dependencies in the general
case, which will lead to \ocaml{Obj.magic}.

To avoid the functional extensionality axiom, we require that all the types used
in our definitions are \emph{setoids}, where we will use \coq{x =e y} to
represent setoid equality. All morphisms need to be \emph{proper} morphisms with respect
to the function \emph{respectfulness} relation. In our development, we use
\begin{coqcode}
Notation "A ~> B" := (* proper morphisms  s.t. forall x y, x =e y -> f x =e f y *).
\end{coqcode}
We add an automatic coercion from \coq{A ~> B} to \coq{A -> B}, provide tactics
to automatically discharge proofs that \coq{forall x y, x =e y -> f x =e f y}
for the simplest cases of \coq{f}, and we provide a library with
definitions for basic building blocks: sum and product types, exponentials,
composition, identity,
\ldots

To avoid axiom K, and extract containers to idiomatic OCaml code, we
define containters in terms of shapes, \emph{base} positions (\coq{APos : Set}),
and \emph{decidable validity predicates} that
specify when a base position is valid in a shape. Since validity predicates are in
\coq{Prop}, they will be erased during extraction, and their decidability
will make unnecessary the use of axiom K.
\begin{coqcode}
Record App F (X : Type) := {shape : Shape F; cont : {p : APos | valid s p = true} -> X}.
\end{coqcode}

Since we require every type to be a setoid, we need to define a setoid equality
(\coq{=e}) for container functors. Given two \coq{x y : App F X}, we define
\coq{x =e y} as the setoid equality of the shapes, and the extensional equality of
the contents, given equal shapes. I.e.\ \coq{x =e y} iff
\begin{coqcode}
shape x =e shape y /\ (forall p p', projT1 p = projT1 p' -> cont x p =e cont y p')
\end{coqcode}

@@

\noindent
where \coq{LFix_in} is the initial algebra $\InIso$, while \coq{LFix_out} is its inverse $\InOp$ (see Section~\ref{sec:rec-schemes}).
As an example, the initial \coq{F}-algebra for the
container that is isomorpic to the functor
\coq{F X = unit + A * X} is the type of lists with the \coq{F}-algebra
being defined by the empty list \coq{Empty : unit -> LFix F} and the cons
operation \coq{Cons : A * LFix F -> LFix F}.

@@

\noindent
We prove that catamorphisms satisfy the universality property we explained in
Section~\ref{sec:rec-schemes}:
\begin{coqcode}
Lemma cata_univ `{eA : setoid A} (alg : Alg F A) (f : LFix ~> A)
    : f =e cata alg <-> f =e alg \o fmap f \o l_out.
\end{coqcode}

@@

As we mentioned in Section~\ref{sec:rec-schemes}, given an $F$-algebra and
$F$-coalgebra, the hylomorphism is the unique solution (when it exists) to the
equation
\[
  f = a \circ F\; f \circ c
\]
As we stated earlier, a solution to this equation does not exist for an
arbitrary  algebra/coalgebra pair and, in fact, this definition cannot be
accepted by Coq.

@@

\subsubsection{On the subtype of finite elements}
As we mention in
Section~\ref{sec:rec-schemes}, we define recursive anamorphisms as
hylomorphisms built by using a recursive coalgebra, and the respective
initial \coq{F}-algebra.
In other words, in this development we have defined recursive anamorphisms on inductive data
types.  We might have as well defined them on the subtype of finite elements of
coinductive data types using a predicate which states when an element of a
coinductive data type is finite:
\begin{coqcode}
Inductive FinF : GFix F -> Prop :=
| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
\end{coqcode}
Now the subtype \coq{{x : GFix F | FinF x}} of finite elements for \coq{GFix F}
is isomorphic its corresponding the inductive data type \coq{LFix F}. This is
easy to see. We first define a catamorphism \coq{ccata_f_} from the
subtype \coq{{x : GFix F | FinF x}} of finitary elements of \coq{GFix F} to
any \coq{F}-algebra.
\begin{coqcode}
Definition ccata_f_ `{eA : setoid A} (g : Alg F A)
  : forall x : GFix F, FinF x -> A := fix f x H :=
    let hx := g_out x in
      g (MkCont (shape hx) (fun e => f (cont hx e) (FinF_inv H e))).
\end{coqcode}
We now prove this is isomorphic to the least fixed-point of the functor
\coq{F}.
We take the catamorphism from the finite elements of \coq{GFix F} to the
inductive data type \coq{LFix F} using the \coq{F}-algebra \coq{l_in}. Its
inverse is the catamorphism on the restriction of \coq{g_in} to the finite
elements of \coq{GFix}, which we denote by \coq{lg_in}.  The following lemmas
prove the isomorphism:
\begin{coqcode}
Lemma cata_ccata `{setoid A} : cata lg_in \o ccata l_in =e id.
Lemma ccata_cata `{setoid A} : ccata l_in \o cata lg_in =e id.
\end{coqcode}
The finite subtype of \coq{GFix F} allows us to compose
catamorphisms and anamorphisms, by using the above isomorphism. In our work,
however, we use \emph{recursive} anamorphisms, defined as follows for
a recursive coalgebra \coq{c}:
\begin{coqcode}
Definition rana c := hylo l_in c.
\end{coqcode}

>>>>

The remainder of the paper is structured as follows:
\begin{itemize}
  \item In Section~\ref{sec:category-theory} we give an overview about recursion
schemes.
  \item In Section~\ref{sec:containers} we formalise the type of container
functors ensuring the presence of least and greatest fixed-points for functors
and suitably adapted for program extraction.
  \item In Section~\ref{sec:recursion-schemes} we mechanise folds, unfolds, and
hylomorphisms, as well as proving their uniqueness properties in Coq.
  \item In Section~\ref{sec:examples} we use the framework to formalise examples
of divide-and-conquer, dynamic programming, and mutual recursion algorithms.
Furthermore we verify the short-cut fusion optimisation and show the extracted
optimised code to OCaml.
\end{itemize}

\section{Recursion Schemes}\label{sec:category-theory}
The structure of data is very similar to the structure of an algorithm which
processes that data. This relationship manifests in the form of structured
recursion schemes which are widely used in functional languages such Haskell.
Canonical examples are folds (catamorphisms), which consume data, and unfolds
(anamorphisms), which produce it. While some implementations of recursion
schemes like $\operator{foldr}$/$\operator{unfoldr}$ in Haskell are specific to
a particular data structure, in this case $\operator{List}$s, we can generalise
further these ideas to account for generic (co)inductive data types and generic
algorithms operating on them.

@@

\subsection{Mechanising Extractable Container Functors}
\label{sec:containers}
Recursion schemes are defined using generic functors representing the scheme
behind the recursive operator.  However, as not all functors possess enough
fixed-points we need to be able to restrict to those mappings \coq{F : Type -> Type}
that are so called \emph{strictly positive}.

A common solution is to consider \emph{containers}~\cite{AbbottAG05}.  A
container $S \triangleright P$ is defined by a type of \emph{shapes} $S : \Type$
and a \emph{family of position types} indexed by shapes $P : S \to \Type$.  An
\emph{extension} of this container is a functor
$\llbracket S \triangleright P \rrbracket$, whose action on objects is given by
$\llbracket S \triangleright P \rrbracket \; X = \Sigma s : S.\; P\;s \to X$,
and on morphisms is given by post-composition, keeping the shape the same
$\llbracket S \triangleright P \rrbracket \; f = \lambda (s,\;g).\;(s,\;f\circ g)$.
Intuitively, the type of shapes captures the different ``cases'' of an algebraic
datatype, and the positions of a shape are used to identify the elements inside
a container of the given shape.

@@

Note however, that the type $\forall s : S.\; P\;s \to X$ would force us to add
the functional extensionality axiom to deal with the heterogeneous equality of
position. Thus, we require that all the types used in our definitions are
\emph{setoids}, where we will use \coq{x =e y} to represent setoid equality. All
morphisms need to be \emph{proper} morphisms with respect to the function
\emph{respectfulness} relation. In our development, we use
\begin{coqcode}
Notation "A ~> B" := (* proper morphisms  s.t. forall x y, x =e y -> f x =e f y *).
\end{coqcode}
We add an automatic coercion from \coq{A ~> B} to \coq{A -> B} and provide tactics
to automatically discharge proofs that \coq{forall x y, x =e y -> f x =e f y}
for the simplest cases of \coq{f}.

We define containers in terms of shapes, \emph{base} positions (\coq{APos : Set}),
and \emph{decidable validity predicates} that
specify when a base position is valid in a shape. Since validity predicates are in
\coq{Prop}, they will be erased during extraction:
\begin{coqcode}
Record App F (X : Type) := {shape : Shape F; cont : {p : APos | valid s p = true} -> X}.
\end{coqcode}

Given two \coq{x y : App F X}, we define
\coq{x =e y} as the setoid equality of the shapes as
\begin{coqcode}
shape x =e shape y /\ (forall p p', projT1 p = projT1 p' -> cont x p =e cont y p')
\end{coqcode}

@@

\noindent
where \coq{LFix_in} is the initial algebra $\InIso$, while \coq{LFix_out} is its inverse $\InOp$ (see Section~\ref{sec:category-theory}).
As an example, the initial \coq{F}-algebra for the
container that is isomorpic to the functor
\coq{F X = unit + A * X} is the type of lists with the \coq{F}-algebra
being defined by the empty list \coq{Empty : unit -> LFix F} and the cons
operation \coq{Cons : A * LFix F -> LFix F}.

@@

\noindent
We prove that catamorphisms satisfy the universality property we explained in
Section~\ref{sec:category-theory}:
\begin{coqcode}
Lemma cata_univ `{eA : setoid A} (alg : Alg F A) (f : LFix ~> A)
    : f =e cata alg <-> f =e alg \o fmap f \o l_out.
\end{coqcode}

@@

As we mentioned in Section~\ref{sec:category-theory}, given an $F$-algebra and
$F$-coalgebra, the hylomorphism is the unique solution (when it exists) to the
equation
\[
  f = a \circ F\; f \circ c
\]
As we stated earlier, a solution to this equation does not exist for an
arbitrary  algebra/coalgebra pair and, in fact, this definition cannot be
accepted by Coq.

@@

\subsubsection{On the subtype of finite elements}
As we mention in
Section~\ref{sec:category-theory}, we define recursive anamorphisms as
hylomorphisms built by using a recursive coalgebra, and the respective
initial \coq{F}-algebra.
In other words, in this development we have defined recursive anamorphisms on inductive data
types.  We might have as well defined them on the subtype of finite elements of
coinductive data types using a predicate which states when an element of a
coinductive data type is finite:
\begin{coqcode}
Inductive FinF : GFix F -> Prop :=
| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
\end{coqcode}
Now the subtype \coq{{x : GFix F | FinF x}} of finite elements for \coq{GFix F}
is isomorphic its corresponding the inductive data type \coq{LFix F}. This is
easy to see. We first define a catamorphism \coq{ccata_f_} from the
subtype \coq{{x : GFix F | FinF x}} of finitary elements of \coq{GFix F} to
any \coq{F}-algebra.
\begin{coqcode}
Definition ccata_f_ `{eA : setoid A} (g : Alg F A)
  : forall x : GFix F, FinF x -> A := fix f x H :=
    let hx := g_out x in
      g (MkCont (shape hx) (fun e => f (cont hx e) (FinF_inv H e))).
\end{coqcode}
We now prove this is isomorphic to the least fixed-point of the functor
\coq{F}.
We take the catamorphism from the finite elements of \coq{GFix F} to the
inductive data type \coq{LFix F} using the \coq{F}-algebra \coq{l_in}. Its
inverse is the catamorphism on the restriction of \coq{g_in} to the finite
elements of \coq{GFix}, which we denote by \coq{lg_in}.  The following lemmas
prove the isomorphism:
\begin{coqcode}
Lemma cata_ccata `{setoid A} : cata lg_in \o ccata l_in =e id.
Lemma ccata_cata `{setoid A} : ccata l_in \o cata lg_in =e id.
\end{coqcode}
The finite subtype of \coq{GFix F} allows us to compose
catamorphisms and anamorphisms, by using the above isomorphism. In our work,
however, we use \emph{recursive} anamorphisms, defined as follows for
a recursive coalgebra \coq{c}:
\begin{coqcode}
Definition rana c := hylo l_in c.
\end{coqcode}
